### 2024-07-02 latex/HP1_comm.tex
new input files

%add comment about yoga replacing mārga and saudha here and in the next two verses.
The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne. However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below). It apears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.

@@

Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}. However, the difference in the sense is so minor that it does alter the meaning.

The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
\emph{-rājamsaudhaṃ}, which he explains as follows:

@@

\begin{enumerate}
\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
    adhirohiṇī}) 
\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The text itself has
    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
    haṭhapradīpikā}) 
\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
\end{enumerate}

According to classical poetology a comparison containing all four elements is termed a ``full
comparison'' (\emph{pūrṇopamā}), whereas an elision of one or more elements is called a ``deficient
comparison'' (\emph{luptopamā}). One element that cannot be omitted without losing the comparison
is the \emph{upamāna}. In the verse as given by Brahmānanda this would be ``the ladder''. The
implication is spelt out by Brahmānanda in his commmentary: ``Just as a ladder leads someone
desiring to ascend to a lofty mansion easily leads him to this mansion, in the same way also the
\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga.'' However, his
text version spells out only the side of the \emph{upameya}, that is, ``Haṭhayoga leads to
Rājayoga'', but leaves the \emph{upamāna} open to interpretation. His mention of the word
\emph{saudha} in the commentary suggests that this was one of the other options he found in
manuscripts—this reading is very well attested—but was rejected by him. In this way Brahmānanda
makes sure that the text states the obvious, but at the same time it loses part of the comparison,
and it loses its dynamics, which is brought out in the reading \emph{rājasaudha} preferred with
good manuscript evidence in our critical edition. We think that Brahmānanda was eager to state at
the outset the primacy of Rājayoga, and to this end sacrificed the more poetically elegant original
reading that mentions the royal mansion as a sort of metaphor for the ``royal yoga''.

@@

As Brahmānanda notes, the compound \emph{kṛpākaraḥ} can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}). In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.       
\end{philcomm}

@@

\begin{philcomm}[hp01_004]   
The word \emph{athavā} (‘or’) is well attested but difficult to construe here. Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it. The variant \emph{mahāyogī} in \epsilonOne and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}. One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.   
\end{philcomm}

@@

Caturbhuja Misra's \emph{Mugdhāvabodhinī} (1.7.8) on the \emph{Rasahṛdayatantra}

@@

The \alphaOne\ and \alphaTwo\ reading of \emph{siddhaḥ kāpālikas} is an exception among the manuscripts and seems too vague to be referring to someone within a lineage. However, Khaṇḍa\-kāpālika is well attested by the manuscripts (including \alphaThree) and this name appears in other literature. Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check??), \emph{Bṛhatkathāmañjarī} 10.45 (check??) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}). It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika. Alternatively, it could simply refer to one who uses a broken skull as a bowl.
\end{philcomm}

%%%%%%%%%%
\subsection*{1.9}
\begin{translation}[hp01_009]
These and other great adepts used the power of hathayoga to smash the rod of death and [so] are roaming the worlds.
\end{translation}

@@

%The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty. 
The \textalpha\ group omits the second hemistich of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated). Both \emph{°jagatām} and \emph{°yuktānām} are well attested by the collated manuscripts. We have adopted \emph{°jagatāṃ} because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds. This reading may not have been understood by some and was changed in other witnesses to \emph{°yuktānām}, which was adopted by Brahmānanda in \emph{Jyotsnā} 1.10.       
\end{philcomm}

@@

\begin{philcomm}[hp01_011]        
Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia, but the manuscript transmission favours the plural.  
\end{philcomm}

%%%%%%%%%%
\subsection*{1.12}
\begin{translation}[hp01_012]
In a well-ruled, righteous region, with plenty of food and free from upheaval, the Haṭhayogi should live in an isolated hut.%JM changed disturbance to upheaval
\end{translation}

\begin{sources}[hp01_012]
\end{sources}

@@

\begin{philcomm}[hp01_012]  
The term \emph{maṭhikā} occurs in narrative literature and yoga texts in the sense of a small hut. For example, in the \textit{Kathāsaritsāgara} (12.9.14, 29–30), \emph{maṭhikā} refers to the small hut built in a cremation ground by a young Brahmin who makes as his bed the ashes of the dead girl he had hoped to marry. In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic. In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four fore-arm lengths (\emph{hasta}) high and wide, and it can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).

In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}. This hemistich derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}. It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water. None of the early manuscripts has this hemistich, which suggests that it was added at a later time. Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition. These manuscripts are not close to an early hyp\-archetype of the text.
\end{philcomm}

%%%%%%%%%%
\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door and is without cracks, holes and bumps. It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way. It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.%JM remove "it is" etc. which read oddly, thus: With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
\end{translation}

@@

\begin{philcomm}[hp01_017]        
The reading of \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group. Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga. The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body. The variant reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and similar formulations occur even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}). It is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., \textbeta, \textgamma\ and \textdelta).
\end{philcomm}

@@

\begin{philcomm}[hp01_021]   
Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (\textalpha\ group) and \getsiglum{V1} (\texteta\ group), the sources and the testimonia. It appears to be verse filler here rather than indicating a temporal sequence of actions. Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first hemistich from a description of this posture in the \emph{Śāradātilakatantra}. This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}. It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).

@@

%%%%%%%%%%
\subsection*{1.22}
\begin{translation}[hp01_022]
Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
\end{translation}

@@

??The word \emph{vyutkrameṇa} appears to describe the position of the ankles. Its basic meaning is ‘against the normal direction,’ which would suggest that the ankles are turned out or crossed rather than placed together naturally. If the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area. See \emph{Yoga Mīmāṃsā}, vol 8, no. 2, pp. 29–30 for a discussion of \emph{vyutkrameṇa} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no. 2, Figures 3–6 for photographs of a practitioner performing this \emph{āsana}. 
\end{philcomm}

%%%%%%%%%%
\subsection*{1.23}
\begin{translation}[hp01_023]
[The yogi] correctly assumes lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
\end{translation}

@@

\begin{philcomm}[hp01_023]
The names \emph{kurkuṭa} and \emph{kurkkuṭa} in some manuscripts are variant spellings of \emph{kukkuṭa} attested in the \emph{Pañcatantra} (M-W).
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.30}
\begin{translation}[hp01_030]
Supporting oneself on the ground with both hands, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.

\begin{sources}[hp01_030]
\emph{Vimānārcanākalpa} 96

@@

\emph{Pādmasaṃhitā} (\emph{yogapāda}) 1.21c–22d:

@@

\emph{Ahirbudhnyasaṃhitā} 31.36–37

@@

\emph{Vasiṣṭhasaṃhitā} 1.76–77

@@

\begin{testimonia}[hp01_030]
\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})

@@

\emph{Haṭharatnāvalī} 3.42

@@

\begin{philcomm}[hp01_030]
%
The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources. The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.

In the second verse quarter, the pronoun in \emph{tatkūrpare} refers to the two hands (\emph{karadvaya}). This is stated more explicitly (i.e., \emph{hastayoḥ kūrparau}) in \emph{Vasiṣṭhasaṃhitā} 1.76c and \emph{Yogayājñavalkya} 3.15c.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.35}
\begin{translation}[hp01_035]
Now, the adept's pose (\emph{siddhāsana}).

[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

\begin{versinnote}
\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
\tl{meḍhre pādam athaikam āsyahṛdaye dhṛtvā samaṃ vigraham |\\+}
\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
\tl{etan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
\tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
\tl{\var{7b meḍhre ] meṃḍhre A • athaikam āsyahṛdaye ] H; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPk, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikaṃ ekahṛdayo T, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBT; kṛtvā GLGPGPkY}\\+}
\tl{\var{7c paśyed ] TH; paśyan VAGHSTvl, paśyad Y • antaraṃ ] VAGPkGLT; antare GBGP}\\+}
\tl{\var{7d hy etan ] H; caitan VAGBGLGPT, etan GPkHS, tv etat Y • °janakaṃ ] °navidhau A, °jananaṃ GB, °nakaraṃ Y • procyate ] idaṃ bhavet Y (unm.)} ?? check\\!}
%?? retain variants for athaikam āsyahṛdaye dhṛtvā samaṃ vigraham
\end{versinnote}

@@

\begin{philcomm}[hp01_035]
%?? Rewrite the note to reflect the adopted reading ekahṛdayo

The adopted reading \emph{ekahṛdayo} is  supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}, which is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, ..

>>>>

%add comment about yoga replacing mārga and saudha here and in the next two verses.
The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne. However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below). It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.

@@

Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
The difference in meaning is negligable.

The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
\emph{-rājasaudhaṃ}, which he explains as follows:

@@

\begin{enumerate}
\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
    adhirohiṇī}) 
\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has
    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
    haṭhapradīpikā}) 
\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
\end{enumerate}

According to classical poetology a comparison containing all four elements is termed a ``full
comparison'' (\emph{pūrṇopamā}), whereas an elision of one or more elements is called a ``deficient
comparison'' (\emph{luptopamā}). One element that cannot be omitted without losing the comparison
is the \emph{upamāna}. In the verse as given by Brahmānanda this would be ``the ladder''. The
implication is spelt out by Brahmānanda in his commmentary: ``Just as a ladder leads someone
desiring to ascend to a lofty mansion easily leads him to this mansion, in the same way also the
\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga.'' However, his
text version spells out only the side of the \emph{upameya}, that is, ``Haṭhayoga leads to
Rājayoga'', but leaves the \emph{upamāna} open to interpretation. His mention of the word
\emph{saudha} in the commentary suggests that this was one of the other options he found in
manuscripts—this reading is very well attested—but was rejected by him. In this way Brahmānanda
makes sure that the text states the obvious, but at the same time it loses part of the comparison,
and it loses its dynamics, which is brought out in the reading \emph{rājasaudha} preferred with
good manuscript evidence in our critical edition. We think that Brahmānanda was eager to state at
the outset the primacy of Rājayoga, and to this end sacrificed the more poetically elegant original
reading that mentions the royal mansion as a metaphor for the ``royal yoga''.

@@

As Brahmānanda notes, the compound \emph{kṛpākaraḥ} can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}). In the Devanāgarī transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.       
\end{philcomm}

@@

\begin{philcomm}[hp01_004]   
The word \emph{athavā} (‘or’) is well attested but difficult to construe here. Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it. The variant \emph{mahāyogī} in \epsilonOne\ and other manuscripts (\getsiglum{G5,J4,J11,Ko}) is probably an attempt to remove the difficulty of understanding \emph{athavā}. One could emend to \emph{tathā} in light of the attested reading \emph{yathā} (\getsiglum{C7}) but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.   
\end{philcomm}

@@

Caturbhuja Miśra's \emph{Mugdhāvabodhinī} (1.7.8) on the \emph{Rasahṛdayatantra}

@@

The \alphaOne\ and \alphaTwo\ reading of \emph{siddhaḥ kāpālikas} is an exception among the manuscripts and seems too vague to be referring to someone within a lineage. Khaṇḍa\-kāpālika is well attested by the remaining manuscripts (including \alphaThree) and this name appears in other texts, e.g.~Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45, where Khaṇḍakāpālika is the first of the 24 Vīras (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}). It is likely to refer to an ascetic who carries a broken skull. \emph{Matsyendrasaṃhitā} 33.2 mentions a practice for which one needs a \emph{khaṇḍakapāla} and in the \emph{Saṃvaramaṇḍala} of the \emph{Niṣpannayogāvalī}, p.26, Vajravārāhī is \emph{kapālakhaṇḍakṛtakaṭibhūṣaṇā}. The compound \emph{khaṇḍakāpālika} is found at \emph{Kathāsaritsāgara} 18.2.6, but there \sl{khaṇḍa} is being used in a derogatory sense (18.2.15 refers to the same character as a \sl{duṣṭakāpālika}).
%It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika. Alternatively, it could simply refer to one who uses a broken skull as a bowl. 
% Reference to the KSS completed and text changed: before: Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check).  Why "part-time".
\end{philcomm}

%%%%%%%%%%
\subsection*{1.9}
\begin{translation}[hp01_009]
These and other great adepts used the power of haṭhayoga to smash the rod of death and [so] are roaming the worlds.
\end{translation}

@@

%The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty. 
The \textalpha\ group omits the second line of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated). Both \emph{°jagatām} and \emph{°yuktānām} are well attested by the collated manuscripts. We have adopted \emph{°jagatāṃ} because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds. This reading may not have been understood by some and was changed in other witnesses to \emph{°yuktānām}, which was adopted by Brahmānanda in \emph{Jyotsnā} 1.10.       
\end{philcomm}

@@

\begin{philcomm}[hp01_011]        
Either the singular or plural of \emph{yogin} could be read here. The singular is well attested among the testimonia, but the manuscript transmission favours the plural.  
\end{philcomm}

%%%%%%%%%%
\subsection*{1.12}
\begin{translation}[hp01_012]
In a well-ruled, righteous region, with plenty of food and free from upheaval, the Haṭhayogi should live in an isolated hut.
\end{translation}

%\begin{sources}[hp01_012]
%\end{sources}

@@

\begin{philcomm}[hp01_012]  
The term \emph{maṭhikā} occurs in narrative literature and yoga texts in the sense of a small hut. For example, in the \textit{Kathāsaritsāgara} (12.9.14, 29–30), \emph{maṭhikā} refers to the small hut built in a cremation ground by a young Brahmin who makes as his bed the ashes of the dead girl he had hoped to marry. In several other stories (\textit{Kathāsaritsāgara} 6.6.132, 10.5.89, 12.25.35), \emph{maṭhikā} is the term used for the hut of an ascetic. In an elaborate description of the huts (\emph{maṭhikā}) used for Haṭhayoga, the author of the \emph{Haṭhābhyāsapaddhati} states that the dimensions of the hut are four "\emph{hasta}s" high and wide (there are various definitions of the term: 18 inches according to the Larger Petrograd Dictionary and Monier-Williams, 48 inches according to the Smaller Petrograd Dictionary; the NWS lists even more variations, but the stipulation that the hut is the length of a bow found in the \emph{Gorakṣaśataka} (on which see below) suggests that 18 inches was meant). The hut can be made of various materials, such as red earth, ashes, plaster and so on (Birch and Singleton 2019: 17–18).

In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional line, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}. This line derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}. It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water. None of the early manuscripts has this line, which suggests that it was added at a later time. Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition. These manuscripts are not close to an early hyp\-archetype of the text.
\end{philcomm}

%%%%%%%%%%
\subsection*{1.13}
\begin{translation}[hp01_013]
With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
%JM changed tr from : It has a small door and is without cracks, holes and bumps. It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way. It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha. 
\end{translation}

@@

\begin{philcomm}[hp01_017]        
The reading \emph{aṅgapāṭavam} is attested among many of the early manuscripts, including the main one of the \textalpha\ group. Although this compound rarely appears in other yoga texts, a similar term \emph{śarīrapāṭava} occurs in the \emph{Śivasaṃhitā} (2.35) as one of the benefits bestowed by digestive fire (\emph{vaiśvānarāgni}), which indicates that the word \emph{pāṭava} was used in relation to the body and the benefits of yoga. The compound \emph{aṅgapāṭava} seems to imply the optimal functioning of the body. The variant reading, \emph{aṅgalāghava} (‘lightness of the limbs’ or ‘dexterity’) is more common in yoga texts and similar formulations occur even in works known to Svātmārāma, such as the \emph{Dattātreyayogaśāstra} (\emph{śarīralaghutā}) and the \textit{Amanaska} ([...] \emph{laghutvaṃ ca śarīrasyopajāyate}). It is likely that the less common term \emph{aṅgapāṭavam} was changed to the more widely used notion of \emph{aṅgalāghava}, perhaps early on in the transmission, as the latter is attested by manuscripts in several early groups (i.e., \textbeta, \textgamma\ and \textdelta).
\end{philcomm}

@@

\begin{philcomm}[hp01_021]   
Although most witnesses have \emph{tathā} in 1.21a, the word \emph{atha} has been accepted because it is attested by \getsiglum{G4} (\textalpha\ group) and \getsiglum{V1} (\texteta\ group), the sources and the testimonia. It appears to be verse filler here rather than indicating a temporal sequence of actions. Svātmārāma borrowed the verse on \emph{vīrāsana} from the \emph{Vasiṣṭhasaṃhitā}, the redactor of which appears to have adapted its first line from a description of this posture in the \emph{Śāradātilakatantra}. This would explain the rather strange syntax of the \emph{Vasiṣṭhasaṃhitā}’s version, in which \emph{adhaḥ kṛtvā} was changed to \emph{athaikasmin}, and \emph{tathetaram} became \emph{ca saṃsthitam}. It seems that \emph{saṃsthitaṃ} must be understood with \emph{ūruṃ} in the third \emph{pāda} in the sense of \emph{saṃsthāpya} (i.e., ‘having placed’).

@@

%%%%%%%%%%
\subsection*{1.22}
\begin{translation}[hp01_022]
Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed. 
\end{translation}

@@

The word \emph{vyutkrameṇa} describes the position of the ankles. Its basic meaning is ‘against the normal direction’. In \emph{āsana} descriptions it usually means ‘crossed’ (see e.g.~\emph{Vasiṣṭhasaṃhitā} 1.71), which is how we have understood it here. It could also mean ‘turned out’: if the yogi is in a kneeling-type position, turning the feet out would bring the ankles together, blocking the perineal area. See \emph{Yoga Mīmāṃsā}, vol 8, no. 2, pp. 29–30 for a discussion of \emph{vyutkrameṇa} and the position of the ankles in \emph{kūrmāsana}, and vol 8, no. 2, figures 3–6 for photographs of a practitioner performing this \emph{āsana}. 
\end{philcomm}

%%%%%%%%%%
\subsection*{1.23}
\begin{translation}[hp01_023]
[The yogi] correctly assumes the lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
\end{translation}
%% JH "the lotus pose" for "lotus pose"

@@

\begin{philcomm}[hp01_023]
The names \emph{kurkuṭa} and \emph{kurkkuṭa} in some manuscripts are variant spellings of \emph{kukkuṭa} attested in the \emph{Pañcatantra} (MW).
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.30}
\begin{translation}[hp01_030]
Supporting oneself on the ground with both palms, the elbows placed on either side of the navel, lifted up into the air in a raised posture [as straight] as a stick: they call this posture the peacock.
%Maybe we should adopt karadvayābhyāṃ in 30a. It is supported by alpha and explains many other variants ending in abhyāṃ, and perhaps its strangeness suggests it was original. JM: yes, but  karasthalābhyāṃ of alpha3 is even better and reflects the °talābhyāṃ of the parallels
\end{translation}

\begin{sources}[hp01_030]
Cf. \emph{Vimānārcanākalpa} 96

@@

Cf. \emph{Pādmasaṃhitā} (\emph{yogapāda}) 1.21c–22d:

@@

Cf. \emph{Ahirbudhnyasaṃhitā} 31.36–37

@@

Cf. \emph{Vasiṣṭhasaṃhitā} 1.76–77

@@

\begin{testimonia}[hp01_030]
\emph{Haṭharatnāvalī} 3.42

\emph{Yogacintāmaṇi} f. 84r (attr. \emph{Haṭhapradīpikā})

@@

@@

\begin{philcomm}[hp01_030]
There is no direct source of this verse, but it has the same elements as two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources. The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.

In the second verse quarter, the pronoun in \emph{tatkūrpare} refers to the two hands. This is stated more explicitly (i.e., \emph{hastayoḥ kūrparau}) in \emph{Vasiṣṭhasaṃhitā} 1.76c and \emph{Yogayājñavalkya} 3.15c.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.35}
\begin{translation}[hp01_035a]
Now, the adept's pose (\emph{siddhāsana}).
\end{translation}

\begin{translation}[hp01_035]
[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

\begin{versinnote}
\tl{yonisthānakam aṅghrimūlaghaṭitaṃ kṛtvā dṛḍhaṃ vinyasen\\+}
\tl{meḍhre pādam athaikam ekahṛdayo dhṛtvā samaṃ vigraham |\\+}
\tl{sthāṇuḥ saṃyamitendriyo 'caladṛśā paśyan bhruvor antaraṃ\\+}
\tl{caitan  mokṣakapāṭabhedajanakaṃ siddhāsanaṃ procyate ||\\+}
%\tl{\var{7a °mūlaghaṭitaṃ ] °mulaghaṭanaṃ A • vinyasen ] GHT; vinyase VA, vinyaset Y}\\+}
\tl{\var{7b athaikaṃ ekahṛdayo ] T; athaikadeśahṛdayo V, athaikam eva niyataṃ AGBGPk, athaikam eva niṣataṃ GL, athaikam eva hṛdayaṃ GP, athaikam ekahṛdayaḥ Y • dhṛtvā ] VGBT; kṛtvā GLGPGPkY}\\+}
\tl{\var{7c paśyan ] VAGHSTvl; paśyed T paśyad Y} \\!}
\end{versinnote}

@@

\begin{philcomm}[hp01_035]
In the second verse quarter, the adopted reading \emph{ekahṛdayo} is supported by two manuscripts of the \emph{Haṭhapradīpikā} (\getsiglum{J2,M1}) and is close to the \textalpha\ reading \emph{ekahṛdaye}. It is also attested by the six-chapter \emph{Viveka\-mārtaṇḍa} and the \emph{Yoga\-cintāmaṇi}, which attributes this verse to an unkown work called the \emph{Pavanayogasaṅgraha}. In this case, \emph{ekahṛdayaḥ} appears to describe the yogi as having his mind focused on one thing. There are many variations of this verse quarter in the \emph{Haṭhapradīpikā} manuscripts, as well as in the manuscripts of the sources and testimonia. Most of the collated witnesses have \emph{athaikam eva hṛdaye dhṛtvā}, which is close to the adopted and \textalpha\ reading. Other readings allude here to the practice of the Jālandhara lock, in which the chin is placed on the chest. This is most clearly seen in the \emph{Jyotsnā}'s version, \emph{hṛdaye kṛtvā hanuṃ susthiram} (`having put the jaw firmly on the chest'). The other well-attested reading, \emph{athaikaṃ eva niyataṃ}, was an attempt to fix the problem of \emph{hṛdaye} by replacing it with \emph{niyatam}, which must be read with \emph{meḍhre pādam athaikaṃ} (`having fixed one foot on the penis'). But \emph{niyatam} is redundant here because of \emph{vinyaset} in the first verse quarter.

%The compound \emph{āsyahṛdaye} is supported by other manuscripts, such as \getsiglum{N10,P1,P6,J16} and the variants \emph{asyahṛdaye} and \emph{asyahṛdayaṃ} also occur (e.g., \getsiglum{J7,J12,A1,V4,V18,V16,P8,P9}).

%MD 2024-5-14: adopt caitan (alpha, beta, and most of the mss of VM) in Pāda d? JM: yes
\end{philcomm}
